### 2025-11-17 cluster-analysis.qmd
20251117 - k-means clustering

There are many approaches to cluster analysis, including model-based clustering, density-based clustering, centroid-based clustering, hierarchical clustering (aka connectivity-based clustering), etc.
In general, cluster analysis intends to maximize intracluster similarities and to minimize intercluster similarities [@Ramasubramanian2016].
Cluster analysis is a form of an [unsupervised approach to machine learning](#sec-machineLearningTypesUnsupervised) in which the groups are not known a priori.
There is a risk of reifying clusters identified in cluster analysis when, in some cases, the clusters may reflect artifacts of the analytic method rather than true distinct groups [@Everitt2011].
In general, the classification should be judged on the extent to which it is *useful* rather than on the extent to which it is true or false [@Everitt2011].

An overview of approaches to cluster analysis in `R` is provided by @Kassambara2017a.
In this chapter, we focus on examples using model-based clustering with the `R` package `mclust` [@R-mclust; @Scrucca2023_packages], which uses Gaussian finite mixture modeling.
Model-based clustering assumes the data are generated by an underlying statistical model, and the goal is to recover the structure of that model.
Model-based clustering represents the data as coming from a mixture of probability distributions.

@@

We can evaluate how many clusters to keep, based on the tradeoff between fit and parsimony, using the `mclust::mclustBIC()` (BIC), `mclust::mclustICL()` (ICL), and `mclust::mclustBootstrapLRT()` (bootstrapped likelihood ratio tests) functions of the `mclust` package [@R-mclust; @Scrucca2023_packages].

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

There are many approaches to cluster analysis, including model-based clustering, density-based clustering, centroid-based clustering (e.g., *k*-means clustering), hierarchical clustering (aka connectivity-based clustering), etc.
In general, cluster analysis intends to maximize intracluster similarities and to minimize intercluster similarities [@Ramasubramanian2016].
Cluster analysis is a form of an [unsupervised approach to machine learning](#sec-machineLearningTypesUnsupervised) in which the groups are not known a priori.
There is a risk of reifying clusters identified in cluster analysis when, in some cases, the clusters may reflect artifacts of the analytic method rather than true distinct groups [@Everitt2011].
In general, the classification should be judged on the extent to which it is *useful* rather than on the extent to which it is true or false [@Everitt2011].

An overview of approaches to cluster analysis in `R` is provided by @Kassambara2017a.
In this chapter, we focus on examples using model-based clustering with the `R` package `mclust` [@R-mclust; @Scrucca2023_packages], which uses Gaussian finite mixture modeling.
Model-based clustering assumes the data are generated by an underlying statistical model, and the goal is to recover the structure of that model.
Model-based clustering represents the data as coming from a mixture of probability distributions.
We also use *k*-means clustering.

@@

We can evaluate how many clusters to keep, based on the tradeoff between fit and parsimony, using the `mclust::mclustBIC()` (BIC), `mclust::mclustICL()` (ICL), and `mclust::mclustBootstrapLRT()` (bootstrapped likelihood ratio tests) functions of the `mclust` package [@R-mclust; @Scrucca2023_packages].
We can also use *k*-means clustering using the `stats::kmeans()` function.

@@

###### Model-Based Clustering {#sec-clusterAnalysisExampleNumTiersQBsModel}

@@

###### *k*-Means Clustering {#sec-clusterAnalysisExampleNumTiersQBsKmeans}

```{r}
tiersQB_kMeans <- kmeans(
  x = player_stats_seasonal_offense_recentQB$fantasyPoints,
  centers = numTiersQB)

tiersQB_kMeans
```

@@

###### Model-Based Clustering {#sec-clusterAnalysisExampleNumTiersRBsModel}

@@

###### *k*-Means Clustering {#sec-clusterAnalysisExampleNumTiersRBsKmeans}

```{r}
tiersRB_kMeans <- kmeans(
  x = player_stats_seasonal_offense_recentRB$fantasyPoints,
  centers = numTiersRB)

tiersRB_kMeans
```

@@

###### Model-Based Clustering {#sec-clusterAnalysisExampleNumTiersWRsModel}

@@

###### *k*-Means Clustering {#sec-clusterAnalysisExampleNumTiersWRsKmeans}

```{r}
tiersWR_kMeans <- kmeans(
  x = player_stats_seasonal_offense_recentWR$fantasyPoints,
  centers = numTiersWR)

tiersWR_kMeans
```

@@

###### *k*-Means Clustering {#sec-clusterAnalysisExampleNumTiersTEsKmeans}

```{r}
tiersTE_kMeans <- kmeans(
  x = player_stats_seasonal_offense_recentTE$fantasyPoints,
  centers = numTiersTE)

tiersTE_kMeans
```

@@

##### Model-Based Clustering  {#sec-clusterAnalysisNumWRtypesModel}

@@

##### *k*-Means Clustering  {#sec-clusterAnalysisNumWRtypesKmeans}

```{r}
wrTypes_kMeans <- kmeans(
  x = merged_dataWRs_cluster %>% select(-gsis_id, -display_name),
  centers = numTypesWR)

wrTypes_kMeans
```
